Time slice analyses

### 2017-11-09 analyses/02-time-slice-analyses.Rmd
Updated pipelined analysis (compiles!)

```{r, eval = FALSE, warning = FALSE}
## Run the function for Beck data
slug <- "BeckLee"

## Ages
out1 <- run.all.disparity.strat(morphospace, tree, type = "Age", FADLAD, inc.nodes = TRUE,
                                bootstraps = 100, metric = c(sum, variances), metric.name = "sum_var")
readr::write_csv(out1, path = paste0("../outputs/age", "_", slug, ".csv"))

## Epochs
out2 <- run.all.disparity.strat(morphospace, tree, type = "Epoch", FADLAD, inc.nodes = TRUE,
                        bootstraps = 100, metric = c(sum, variances), metric.name = "sum_var")
readr::write_csv(out2, path = paste0("../outputs/epoch", "_", slug, ".csv"))
```

We also ran the models using varying numbers of bins as follows.

```{r, eval = FALSE}
## Run the function for Beck data
slug <- "BeckLee"

@@

<!-- TG: Suggestion for a "testing the effect of binning/slicing" pipeline -->

@@

Plots
Statistics

>>>>

> TG: for the following pipeline, I've reduced the number of rarefaction that highly increased the calculation time.
We now only calculate the disparity for the non-rarefied subsamples and the subsamples rarefied to contain only 3 taxa (the lowest level of rarefaction).
We can use the full rarefaction if we want to plot rarefaction curves later on.

```{r, eval = FALSE, echo = FALSE}
## Testing data: TO REMOVE
#TG: it's just faster this way
#TG: also, I've only tested it on "Epoch" since it take ages to run, should be pipelined all right though
data(BeckLee_mat99)
data(BeckLee_tree)
data(BeckLee_ages)

morphospace <- BeckLee_mat99
tree <- BeckLee_tree
FADLAD <- BeckLee_ages

```{r, eval = TRUE, warning = FALSE}
## Run the function for Beck data
slug <- "BeckLee"

## Ages (rarefied to 3 taxa only)
out1 <- run.all.disparity.wrapper(morphospace, tree, type = "Age", FADLAD, inc.nodes = TRUE,
                                bootstraps = 100, metric = c(sum, variances),
                                metric.name = "sum_var", rarefaction = FALSE)
#readr::write_csv(out1, path = paste0("../outputs/age", "_", slug, ".csv"))

## Epochs (rarefied to 3 taxa only)
out2 <- run.all.disparity.wrapper(morphospace, tree, type = "Epoch", FADLAD, inc.nodes = TRUE,
                        bootstraps = 100, metric = c(sum, variances), metric.name = "sum_var",
                        rarefaction = FALSE)
#readr::write_csv(out2, path = paste0("../outputs/epoch", "_", slug, ".csv"))
```

Note that you can use the previous output format (the data tables) as before by accessing them through `out$results$stratigraphy/duration/number`.
Writing them as a csv will require some thinking since they don't have the same number of columns (though it's still doable I guess).
Also, if you want to plot things through `ggplot`, you can use the values in these data tables!

<!-- TG: I've not run this analysis below -->
<!-- We also ran the models using varying numbers of bins as follows. -->

```{r, eval = FALSE, echo = FALSE}
## Run the function for Beck data
slug <- "BeckLee"

@@

<!-- NC: Need to think more about the outputs here, currently they don't make much sense as there are no column headers -->

@@

The results as a table

```{r, eval = TRUE, echo = FALSE}
knitr::kable(out2$results$stratigraphy, caption = "Disparity per stratigraphic time: unequal time bins and non-equidistant time slices")
knitr::kable(out2$results$duration, caption = "Disparity per average stratigraphic time: equal time bins and equidistant time slices based on the median stratigraphic epoch duration")
knitr::kable(out2$results$number, caption = "Disparity per average stratigraphic time: equal time bins and equidistant time slices based on the number of stratigraphic epochs")
```

This is not really digest and we should find a better way to present that.
Anyway, these boring tables are always good for the supplementaries.

## Plotting the results `dispRity` style

Here is an example of the results using `plot.dispRity`.
The upper panel is with the CIs and the lower one is the same without the CIs.

### The results for the Age type

```{r, eval = TRUE, fig.width = 12, fig.height = 18, echo = TRUE}
op <- par(bty = "n", mfrow = c(3,2))
## Principal colors
colors <- palette()[1:5][-1] #TG: that can be made fancier
colors[5] <- "black"
## Color combination (for the CI)
colors_CI <- c("pink", "lightgreen", "lightblue", "lightcyan", "lightgray")
## y limits
ylim <- range(unlist(lapply(out1$results, function(X) return(c(min(X[,5], na.rm = TRUE), max(X[,5],
  na.rm = TRUE))))))
## Zooming
ylim_zoom <- c(20, 35)

## Plotting the different methods (with/without CIs)
plot.results.dispRity(out1$object, method = 1, colors = colors, colors_CI = colors_CI,
  main = "Stratigraphy", ylim = ylim, CI = TRUE, legend = TRUE)
plot.results.dispRity(out1$object, method = 1, colors = colors, colors_CI = colors_CI,
  main = "Stratigraphy (no CI)", ylim = ylim_zoom, CI = FALSE, time.bins.line = TRUE)

plot.results.dispRity(out1$object, method = 2, colors = colors, colors_CI = colors_CI,
  main = "Duration", ylim = ylim, CI = TRUE)
plot.results.dispRity(out1$object, method = 2, colors = colors, colors_CI = colors_CI,
  main = "Duration (no CI)", ylim = ylim_zoom, CI = FALSE, time.bins.line = TRUE)

plot.results.dispRity(out1$object, method = 3, colors = colors, colors_CI = colors_CI,
  main = "Number", ylim = ylim, CI = TRUE)
plot.results.dispRity(out1$object, method = 3, colors = colors, colors_CI = colors_CI,
  main = "Number (no CI)", ylim = ylim_zoom, CI = FALSE, time.bins.line = TRUE)
```

### The results for the Epoch type

```{r, eval = TRUE, fig.width = 12, fig.height = 18, echo = TRUE}
op <- par(bty = "n", mfrow = c(3,2))
## Principal colors
colors <- palette()[1:5][-1] #TG: that can be made fancier
colors[5] <- "black"
## Color combination (for the CI)
colors_CI <- c("pink", "lightgreen", "lightblue", "lightcyan", "lightgray")
## y limits
ylim <- range(unlist(lapply(out2$results, function(X) return(c(min(X[,5], na.rm = TRUE),
  max(X[,5], na.rm = TRUE))))))

## Zooming
ylim_zoom <- c(20, 35)

## Plotting the different methods (with/without CIS)
plot.results.dispRity(out2$object, method = 1, colors = colors, colors_CI = colors_CI,
  main = "Stratigraphy", ylim = ylim, CI = TRUE, legend = TRUE)
plot.results.dispRity(out2$object, method = 1, colors = colors, colors_CI = colors_CI,
  main = "Stratigraphy (no CI)", ylim = ylim_zoom, CI = FALSE, time.bins.line = TRUE)

plot.results.dispRity(out2$object, method = 2, colors = colors, colors_CI = colors_CI,
  main = "Duration", ylim = ylim, CI = TRUE)
plot.results.dispRity(out2$object, method = 2, colors = colors, colors_CI = colors_CI,
  main = "Duration (no CI)", ylim = ylim_zoom, CI = FALSE, time.bins.line = TRUE)

plot.results.dispRity(out2$object, method = 3, colors = colors, colors_CI = colors_CI,
  main = "Number", ylim = ylim, CI = TRUE)
plot.results.dispRity(out2$object, method = 3, colors = colors, colors_CI = colors_CI,
  main = "Number(no CI)", ylim = ylim_zoom, CI = FALSE, time.bins.line = TRUE)

## Plotting the results histogram style

Another way for the plots would be kind of our histogram representation (you can certainly use `ggplot` here, below is just an example):

### The results per method

```{r, fig.width = 12, fig.height = 18}
## Set the list of available models
models <- list("equalbins", "acctran", "deltran", "punctuated", "gradual")

## Get the stratigraphic, duration and number data for the ages
strat_data_ages <- lapply(models, make.class.histogram, out1$results, method = 1)
durat_data_ages <- lapply(models, make.class.histogram, out1$results, method = 2)
numbe_data_ages <- lapply(models, make.class.histogram, out1$results, method = 3)

## Get the stratigraphic, duration and number data for the epochs
strat_data_epoc <- lapply(models, make.class.histogram, out2$results, method = 1)
durat_data_epoc <- lapply(models, make.class.histogram, out2$results, method = 2)
numbe_data_epoc <- lapply(models, make.class.histogram, out2$results, method = 3)

## The lower age and the lower disparity limit are arbitrarily fixed to 0 and 15 respectively.
lower_age <- 0; lower_lim <- 15
#TG: might want to change that for other datasets!

## limits functions
get.break <- function(x) return(x$breaks)
get.count <- function(x) return(x$counts)
## Ages limits
xlim_ages <- c(max(unlist(c(sapply(strat_data_ages, get.break), sapply(durat_data_ages, get.break),
  sapply(numbe_data_ages, get.break)))), lower_age)
ylim_ages <- c(lower_lim, max(unlist(c(sapply(strat_data_ages, get.count),
  sapply(durat_data_ages, get.count), sapply(numbe_data_ages, get.count)))))
## Epoch limits
xlim_epoc <- c(max(unlist(c(sapply(strat_data_epoc, get.break), sapply(durat_data_epoc, get.break),
  sapply(numbe_data_epoc, get.break)))), lower_age)
ylim_epoc <- c(lower_lim, max(unlist(c(sapply(strat_data_epoc, get.count),
  sapply(durat_data_epoc, get.count), sapply(numbe_data_epoc, get.count)))))

## Colors
colors <- palette()[1:5]
colors[1] <- "grey"

## Graphics
op <- par(bty = "n", mfrow = c(3,2))

## Plot the stratigraphy - ages
plot(strat_data_ages[[1]], xlim = xlim_ages, ylim = ylim_ages , xlab = "Time (Mya)",
  ylab = "Sum of variance", col = colors[1], main = "Stratigraphy - ages")
## Plot the slicing
for(slice in 2:5) {
  lines(strat_data_ages[[slice]]$breaks, strat_data_ages[[slice]]$counts, col = colors[slice])
}

## Adding the legend
legend(x = "bottomleft", legend = c("time bins", "acctran", "deltran", "punctuated", "gradual"),
  col = colors[1:5], lty = 1, bg = "white")

## Plot the stratigraphy - epoch
plot(strat_data_epoc[[1]], xlim = xlim_epoc, ylim = ylim_epoc , xlab = "Time (Mya)",
  ylab = "Sum of variance", col = colors[1], main = "Stratigraphy - Epoch")
## Plot the slicing
for(slice in 2:5) {
  lines(strat_data_epoc[[slice]]$breaks, strat_data_epoc[[slice]]$counts, col = colors[slice])
}

## Plot the duration - ages
plot(durat_data_ages[[1]], xlim = xlim_ages, ylim = ylim_ages , xlab = "Time (Mya)",
  ylab = "Sum of variance", col = colors[1], main = "Duration - ages")
## Plot the slicing
for(slice in 2:5) {
  lines(durat_data_ages[[slice]]$breaks, durat_data_ages[[slice]]$counts, col = colors[slice])
}

## Plot the duration - epoch
plot(durat_data_epoc[[1]], xlim = xlim_epoc, ylim = ylim_epoc , xlab = "Time (Mya)",
  ylab = "Sum of variance", col = colors[1], main = "Duration - Epoch")
## Plot the slicing
for(slice in 2:5) {
  lines(durat_data_epoc[[slice]]$breaks, durat_data_epoc[[slice]]$counts, col = colors[slice])
}

## Plot the number - ages
plot(numbe_data_ages[[1]], xlim = xlim_ages, ylim = ylim_ages , xlab = "Time (Mya)",
  ylab = "Sum of variance", col = colors[1], main = "Number - ages")
## Plot the slicing
for(slice in 2:5) {
  lines(numbe_data_ages[[slice]]$breaks, numbe_data_ages[[slice]]$counts, col = colors[slice])
}

## Plot the number - epoch
plot(numbe_data_epoc[[1]], xlim = xlim_epoc, ylim = ylim_epoc , xlab = "Time (Mya)",
  ylab = "Sum of variance", col = colors[1], main = "Number - Epoch")
## Plot the slicing
for(slice in 2:5) {
  lines(numbe_data_epoc[[slice]]$breaks, numbe_data_epoc[[slice]]$counts, col = colors[slice])
}
```

Some of the plots are a bit off set and everything but it gives us an idea

### The results per model

Something like that but with transparency:

```{r, fig.width = 6, fig.height = 6}
plot(strat_data_epoc[[1]], xlim = xlim_epoc, ylim = ylim_epoc , xlab = "Time (Mya)",
  ylab = "Sum of variance", col = colors[1], main = "Epoch")
plot(durat_data_epoc[[1]], xlim = xlim_epoc, ylim = ylim_epoc , xlab = "Time (Mya)",
  ylab = "Sum of variance", col = colors[2], main = "", add = TRUE)
plot(numbe_data_epoc[[1]], xlim = xlim_epoc, ylim = ylim_epoc , xlab = "Time (Mya)",
  ylab = "Sum of variance", col = colors[3], main = "", add = TRUE)
```
